GMSE: an R package for generalised management strategy evaluation

### 2017-08-09 notebook/ms/ms.Rmd
Minor tweaks

To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions [@Keith2011]. This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [@Bunnefeld2011; @Bunnefeld2014]. Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016]. Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on simple rules, and user behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017]. Here we introduce generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.

@@

Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders). Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step. Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017]. In finding strategies, GMSE models this trade-off by setting a fixed budget for managers and users (Table 1). Allocations from a manager's budget can be used to increase the cost it takes a user to perform an action, and allocations from a user's budget can be used to perform the action at the cost set by the manager. Hence, stakeholders can have incomplete control over resource use and express competing management objectives.

@@

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops [e.g., @Madsen2017]. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. We consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size. Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for 10 farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

```{r, echo = TRUE}
sim <- gmse(land_ownership = TRUE, stakeholders = 10, observe_type = 0, 
            res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 1000, 
            user_budget = 5000, manager_budget = 5000, res_consume = 1, 
            scaring = TRUE, times_observe = 1, agent_view = 10, 
            plotting = FALSE);
```

>>>>

To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions [@Keith2011]. This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [@Bunnefeld2011; @Bunnefeld2014]. Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016]. Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on fixed rules, and user behaviour likewise remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017]. Here we introduce generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.

@@

Consistent with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for agents (stakeholders). Instead, genetic algorithms are used to heuristically find an adaptive strategy for each stakeholder in each time step. Critically, all stakeholders involved in resource conservation are constrained in their decision-making; managing and using resources takes effort (e.g., time or money), and effort expended in developing or enforcing one policy (for managers) or performing one action (for users) will be effort not expendable elsewhere [@Milner-Gulland2011; @Muller-Hansen2017; @Schluter2017]. In finding strategies, GMSE models this trade-off by setting a fixed budget for managers and users. Allocations from a manager's budget can be used to increase the cost it takes a user to perform an action (i.e., 'policy'), and allocations from a user's budget can be used to perform the action at the cost set by the manager. Hence, stakeholders can have incomplete control over resource use and express competing management objectives.

@@

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops [e.g., @Madsen2017]. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. We consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size. Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for 10 farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

```{r, echo = TRUE}
sim <- gmse(land_ownership = TRUE, stakeholders = 10, observe_type = 0, 
            res_death_K = 2000, manage_target = 1000, RESOURCE_ini = 1000, 
            user_budget = 5000, manager_budget = 5000, res_consume = 1, 
            scaring = TRUE, times_observe = 1, plotting = FALSE);
```
